A Memory-Bounded, Deterministic and Terminating Semantics for the Synchronous Programming Language~\CEU

### 2018-04-16 03-semantics.tex
Continue refactoring of Section 3

The~$\ceu{\Mem(id)}$ primitive represents an assignment or system
call that affects the memory location identified by~$id$.\footnotemark\
%
Following the synchronous hypothesis, $\ceu{\Mem}$ statements and
expressions are considered to be atomic and instantaneous.
%
% As the challenging parts of \CEU reside on its control structures, we are not
% concerned here with a precise semantics for side effects, but only with their
% occurrences in programs.
%
%The special notation $nop$ is used to represent an innocuous $mem$ expression
%(it can be thought as a synonym for $mem(\epsilon)$, where $\epsilon$ is an
%unused identifier).
%
\footnotetext{Although the same symbol~$\ceu{\Id}$ is used in their
  definition, $\ceu{\Mem}$, $\ceu{\AwaitExt}$, $\ceu{\AwaitInt}$
  and~$\ceu{\EmitInt}$ do not share identifiers: any identifier is either a
  variable, an external event, or an internal event.}

Most statements in the abstract language are mapped to their counterparts in
the concrete language.  The exceptions are the finalization
block~$\ceu{\Fin{p}}$ and the \texttt{@}-statements which do not exist in
the concrete language.  (These result from the expansion of the transition
rules to be presented.)

A further difference between the concrete and abstract languages regards the
emit-await pair.  In the concrete language, the \code{await} can be used as
an expression which evaluates to the value stored in the emitted
event, e.g., an
``\code{emit a(10)}'' awakes a ``\code{v=await a}'' setting
variable~\code{v} to~10.  Although the abstract $\ceu{\AwaitInt}$
and~$\ceu{\EmitInt}$ do not support such communication of values, it can be
easily simulated: one can use a shared variable to hold the value of an
$\ceu{\EmitInt}$ and access it after the corresponding $\ceu{\AwaitInt}$
awakes.

@@

\subsection{Operational Semantics}

The operational semantics is a mathematical model that describes how an
abstract \CEU\ programs reacts to a single external input event, i.e., how
starting from this input event the program advances its state until all its
trails are blocked waiting for a another input event.
%%
The formalism we use here is that of small-step operational
semantics~\cite{Plotkin-G-D-1981} in which the meaning of programs is
defined in terms of transitions of an abstract machine.  Each transition
transforms a triple consisting of a program~$p$, a stack level~$n$, and an
emitted event~$e$ into a modified triple as follows:
\[
  \<p,n,e>\trans\<p',n',e'>\,,
\]
where~$p,p'\in\P$ are abstract-language programs, $n,n'\in\N$ are non-negative
integers representing the current stack level, and~$e,e'\in\E\cup\{\nil\}$ are
the events emitted before and after the transition (both possibly being the
empty event~$\nil$).
%$\E$ is a set of primitive events.

@@

At the beginning of a reaction to an input event~$id$, the input description is
initialized with stack level~0 ($n=0$) and with the externally emitted event~$e=id$.
%, but \code{emitInt} expressions can increase the stack level.
At the end of a reaction, after an arbitrary but finite number of transitions,
the last output description will block with a possibly modified program~$p'$ at stack
level~0 and with no event emitted:
\[
  \<p,0,e>\mathbin{\trans[*]}\<p',0,\nil>\,.
\]

@@

Rule \R{push} can be applied whenever there is a nonempty event in the input
description,
and instantly broadcasts the event to the program, which means
    (a)~awaking any active $\ceu{\AwaitExt}$ or $\ceu{\AwaitInt}$ statements
    (see $\bcast$ in
        Figure~\ref{fig.bcast}),
    (b)~creating a nested reaction by increasing the stack level, and, at the same time,
    (c)~consuming the event ($e$ becomes~$\nil$).
%
Rule \R{push} is the only rule that matches an
emitted event and also immediately consumes it.

@@

Given that our semantics focuses on control, rules \R{if-true} and
\R{if-false} are the only to query~$\ceu{\Mem}$ as expressions.
%
Function~$\eval$ evaluates a~$\ceu{\Mem}$ expression to a boolean value.
%
%Although the value here is arbitrary, it is unique in a reaction, because a
%given expression can execute only once within it (remember that $loops$ must
%contain $awaits$ which, from rule \textbf{await}, cannot awake in the same
%reaction they are reached).
%For all other rules, we omit these values (e.g., \textbf{seq-nop}).

@@

A reaction eventually blocks in~$\ceu{\AwaitExt}$, $\ceu{\AwaitInt}$,
$\ceu{\Every}$, $\ceu{\Fin}$, and~$\ceu{\CanRun}$ statements in parallel
trails.
%
Then, if none of the trails is blocked in~$\ceu{\CanRun}$ statements, it means
that the program cannot advance in the current reaction.
%
However, $\ceu{\CanRun}$ statements can still resume at lower stack indexes
and will eventually resume in the current reaction (see rule \R{pop}).

@@

\subsection{Properties}

@@

Transitions~$\out$ and~$\nst$ are defined in such a way that given an input
description either no rule is applicable or exactly one of them can be
applied.  This means that the resulting transition relation~$\trans$ is in
fact a partial function.

The next two lemmas establish the determinism of a single application
of~$\out$ and~$\nst$.  Lemma~\ref{lem.x.det-out} follows from a simple
inspection of rules~\R{push} and~\R{pop}.  The proof of
Lemma~\ref{lem.x.det-nst} follows by induction on the structure
of the derivation trees produced by the rules for~$\nst$.  Both lemmas are
used in the proof of the next theorem.

@@

The main result of this section, Theorem~\ref{thm.x.det}, establishes that
any given number~$i\ge0$ of applications of arbitrary transition rules,
starting from the same input description, will always lead to the same output
description.  In other words, any finite sequence of transitions behave
deterministically.

@@

\begin{definition}
  \label{def.x.Hnst}
  %%
  A description~$\delta=\<p,n,e>$ is \emph{nested-irre\-ducible}
  iff~$e\ne\nil$ or~$p=\ceu{\Nop}$ or~$p=\ceu{\Break}$
  or~$\isblocked(p,n)$.\footnote{We often abbreviate ``$p=\ceu{\Nop}$
    or~$p=\ceu{\Break}$'' as~``$p=\ceu{\Nop,\Break}$''.}
\end{definition}

@@

The syntactic restriction discussed in Section~\ref{sec.ceu.evts}, regarding
the body of loops, and the restriction mentioned earlier in this section,
regarding the body of~$\ceu{\Fin}$ statements, are formalized in
Assumption~\ref{ass.x.syn-rest} below.  These restrictions are essential to
prove the next theorem, Theorem~\ref{thm.x.term-nst-*}.

\begin{assumption}[Syntactic restrictions]\strut
  \label{ass.x.syn-rest}
  %%
  \begin{enumerate:a}
  \item\label{ass.x.syn-rest.fin} If~$p=\ceu{\Fin{p_1}}$ then~$p_1$ contains
    no occurrences of~$\ceu{\AwaitExt}$, $\ceu{\AwaitInt}$, $\ceu{\Every}$,
    or~$\ceu{\Fin}$ expressions.  Consequently, for any~$n$,
    $\<\clear(p_1),n,\nil>\nst[*]\<\ceu{\Nop},n,\nil>$.
    %%
    % \[
    %   \<\clear(p_1),n,\nil>\nst[*]\<\ceu{\Nop},n,\nil>\,.
    % \]
    %%
  \item\label{ass.x.syn-rest.loop} If~$p=\ceu{\Loop{p_1}}$ then all
    execution paths of~$p_1$ contain a matching~$\ceu{\Break}$ or
    an~$\ceu{\AwaitExt}$.  Consequently, for all~$n$, there are~$p_1'$
    and~$e$ such that
    $\<\ceu{\Loop{p_1}},n,\nil>\nst[*]\<p_1',n,e>$,
    %%
    % \[
    %   \<\ceu{\Loop{p_1}},n,\nil>\nst[*]\<p_1',n,e>\,,
    % \]
    %%
    where $p_1'=\ceu{{\Break}\AtLoop{p_1}}$ or~$\isblocked(p_1',n)$\,.
  \end{enumerate:a}
\end{assumption}

@@

The main result of this section, Theorem~\ref{thm.x.term}, is similar to
Theorem~\ref{thm.x.term-nst-*} but applies to transitions~$\trans$ in
general.  Before stating and proving it, we need to characterize irreducible
descriptions in general.  This characterization depends on the notions of
potency and rank.

\begin{definition}
  \label{def.x.pot}
  %%
  The \emph{potency} of a program~$p$ in reaction to event~$e$,
  denoted~$\pot(p,e)$, is the maximum number of~$\ceu{\EmitInt}$ expressions
  that can be executed during a reaction of~$p$ to~$e$.

  More formally, $\pot(p,e)=\pot'(\bcast(p,e))$ where~$\pot'$ is an
  auxiliary function that counts the maximum number of
  reachable~$\ceu{\EmitInt}$ expressions in the program resulting from the
  broadcast of event~$e$ to~$p$.

  The auxiliary function~$\pot'$ is defined by
  clauses~\eqref{def.x.pot.first}--\eqref{def.x.pot.last}:
  \begin{enumerate:a}
  \item\label{def.x.pot.first}$\pot'(\ceu{\EmitInt}(e))=1$.
    %%
  \item$\pot'(\ceu{\IfElse{\Mem(\Id)}{p_1}{p_2}})
    =\max\{\pot'(p_1),\pot'(p_2)\}$.
    %%
  \item$\pot'(\ceu{\Loop{p_1}})=\pot'(p_1)$.
    %%
  \item$\pot'(\ceu{{p_1}\And{p_2}})=\pot'(p_1)+\pot'(p_2)$.
    %%
  \item$\pot'(\ceu{{p_1}\Or{p_2}})=\pot'(p_1)+\pot'(p_2)$.
    %%
  \item If~$p_1\ne\ceu{\Break},\ceu{\AwaitExt(e)}$,
    \begin{align*}
      \pot'(\ceu{p_1;\,p_2})&=\pot'(p_1)+\pot'(p_2)\\
      \pot'(\ceu{p_1\AtLoop{p_2}})&=
      \begin{cases}
        \pot'(p_1)              &\text{if\enspace(\dag)}\\
        \pot'(p_1)+\pot'(p_2)   &\text{otherwise,}\\
      \end{cases}
    \end{align*}
    where~(\dag) stands for: ``a~$\ceu{\Break}$ or~$\ceu{\AwaitExt}$ occurs
    in all execution paths of~$p_1$''.
    %%
  \item If~$p_1,p_2\ne\ceu{\Break}$.
    \[
      \pot'(\ceu{{p_1}\AtAnd{p_2}})=\pot'(p_1)+\pot'(p_2)\,.
    \]
    %%
  \item\label{def.x.pot.before-last} If~$p_1,p_2\ne\ceu{\Break}$
    and~$p_1,p_2\ne\ceu{\Nop}$,
    \[
      \pot'(\ceu{{p_1}\AtOr{p_2}})=\pot'(p_1)+\pot'(p_2)\,.
    \]
    %%
  \item\label{def.x.pot.last} Otherwise, if none
    of~\eqref{def.x.pot.first}--\eqref{def.x.pot.before-last} applies,
    $\pot(\_)=0$.
  \end{enumerate:a}
\end{definition}

@@

@@

The next two lemmas establish that a single application of~$\out$ or~$\nst$
either preserves or decreases the rank of the input description.  All rank
comparisons assume lexicographic order, i.e., if~$\rank(\delta)=\<i,j>$
and~$\rank(\delta')=\<i',j'>$ then~$\rank(\delta)>\rank(\delta')$ iff~$i>i'$
or~$i=i$ and~$j>j'$.
%%
The proof of Lemma~\ref{lem.x.rank-out} follows directly from~\R{push}
and~\R{pop} and from Definitions~\ref{def.x.pot} and~\ref{def.x.rank}.  The
proof of Lemma~\ref{lem.x.rank-nst}, however, is by induction on the
structure of~$\nst$ derivations.  Since Lemma~\ref{lem.x.rank-nst} is of
fundamental importance for the termination result, we detail the proof of
one of its cases.

>>>>

The~$\ceu{\Mem(id)}$ primitive represents a read or write to the memory
location identified by~$id$.\footnotemark\
%
Following the synchronous hypothesis, $\ceu{\Mem}$ statements and
expressions are considered to be atomic and instantaneous.
%
% As the challenging parts of \CEU reside on its control structures, we are not
% concerned here with a precise semantics for side effects, but only with their
% occurrences in programs.
%
%The special notation $nop$ is used to represent an innocuous $mem$ expression
%(it can be thought as a synonym for $mem(\epsilon)$, where $\epsilon$ is an
%unused identifier).
%
\footnotetext{Although the same symbol~$\ceu{\Id}$ is used in their
  definition, $\ceu{\Mem}$, $\ceu{\AwaitExt}$, $\ceu{\AwaitInt}$
  and~$\ceu{\EmitInt}$ do not share identifiers: any identifier is either a
  variable, an external event, or an internal event.}

Most statements in the abstract language are mapped to their counterparts in
the concrete language.  The exceptions are the finalization
block~$\ceu{\Fin{p}}$ and the \texttt{@}-statements which do not exist in
the concrete language.  These result from the expansion of the transition
rules to be presented.

A further difference between the concrete and abstract languages regards the
emit-await pair.  In the concrete language, the \code{await} can be used as
an expression which evaluates to the value stored in the corresponding emitted
event, e.g., an
``\code{emit a(10)}'' awakes a ``\code{v=await a}'' setting
variable~\code{v} to~10.  Although the abstract $\ceu{\AwaitInt}$
and~$\ceu{\EmitInt}$ do not support such communication of values, it can be
easily simulated: one can use a shared variable to hold the value of an
$\ceu{\EmitInt}$ and access it after the corresponding $\ceu{\AwaitInt}$
awakes.

@@

\subsection{Operational Semantics}
\label{sec.sem.opsem}

The operational semantics is a mathematical model that describes how an
abstract \CEU\ programs reacts to a single external input event, i.e., how
starting from this input event the program advances its state until all its
trails are blocked waiting for a another input event.
%%
The formalism we use here is that of small-step operational
semantics~\cite{Plotkin-G-D-1981} in which the meaning of programs is
defined in terms of transitions of an abstract machine.  Each transition
transforms a triple consisting of a program~$p$, a stack level~$n$, and an
emitted event~$e$ into a possibly different triple, i.e.,
\[
  \<p,n,e>\trans\<p',n',e'>\,,
\]
where~$p,p'\in\P$ are abstract-language programs, $n,n'\in\N$ are non-negative
integers representing the current stack level, and~$e,e'\in\E\cup\{\nil\}$ are
the events emitted before and after the transition (both possibly being the
empty event~$\nil$).
%$\E$ is a set of primitive events.

@@

At the beginning of a reaction to an input event~$id$, the input description is
initialized with stack level~0 ($n=0$) and with the externally emitted event~$e=id$.
%, but \code{emitInt} expressions can increase the stack level.
At the end of a reaction, after an arbitrary but finite number of transitions,
the last output description will block with a possibly modified program~$p'$ at stack
level~0 and no event emitted:
\[
  \<p,0,e>\mathbin{\trans[*]}\<p',0,\nil>\,.
\]

@@

Rule \R{push} can be applied whenever there is a nonempty event in the input
description,
and instantly broadcasts the event to the program, which means
    (i)~awaking any active $\ceu{\AwaitExt}$ or $\ceu{\AwaitInt}$ statements
    (see $\bcast$ in
        Figure~\ref{fig.bcast}),
    (ii)~creating a nested reaction by increasing the stack level, and, at the same time,
    (iii)~consuming the event ($e$ becomes~$\nil$).
%
Rule \R{push} is the only rule that matches an
emitted event and also immediately consumes it.

@@

Rules \R{if-true} and \R{if-false} are the only rules that use~$\ceu{\Mem}$
in a way that affects the control flow.
%
Function~$\eval$ evaluates a~$\ceu{\Mem}$ expression to a boolean value.
%
%Although the value here is arbitrary, it is unique in a reaction, because a
%given expression can execute only once within it (remember that $loops$ must
%contain $awaits$ which, from rule \textbf{await}, cannot awake in the same
%reaction they are reached).
%For all other rules, we omit these values (e.g., \textbf{seq-nop}).

@@

A reaction eventually blocks in~$\ceu{\AwaitExt}$, $\ceu{\AwaitInt}$,
$\ceu{\Every}$, $\ceu{\Fin}$, and~$\ceu{\CanRun}$ statements in parallel
trails.
%
Then, if none of the trails is blocked in~$\ceu{\CanRun}$, it means that the
program cannot advance in the current reaction.
%
However, $\ceu{\CanRun}$ statements can still resume at lower stack indexes
and will eventually resume in the current reaction (see rule \R{pop}).

@@

\subsection{Properties}
\label{sec.sem.props}

@@

Transitions~$\out$ and~$\nst$ are defined in such a way that given an input
description either no rule is applicable or exactly one of them can be
applied (no choice involved).  This coupled with the fact that the output of
every rule is a function of its input implies that transitions are
deterministic: the same input description, if it can transition, will always
result in the same output description.  In other words, the transition
relation~$\trans$ is in fact a partial function.

The next two lemmas establish the determinism of a single application
of~$\out$ and~$\nst$.  Lemma~\ref{lem.x.det-out} follows from a simple
inspection of rules~\R{push} and~\R{pop}.  The proof of
Lemma~\ref{lem.x.det-nst} follows by induction on the structure of the
derivation trees produced by the rules for~$\nst$.  Both lemmas are used in
the proof of the Theorem~\ref{thm.x.det}.

@@

The main result of this section, Theorem~\ref{thm.x.det}, establishes that
any given number~$i\ge0$ of applications of arbitrary transition rules,
starting from the same input description, will always lead to the same
output description, i.e., any finite sequence of transitions behave
deterministically.

@@

\begin{definition}
  \label{def.x.Hnst}
  %%
  A description~$\delta=\<p,n,e>$ is \emph{nested-irre\-ducible}
  iff~$e\ne\nil$ or~$p=\ceu{\Nop}$ or~$p=\ceu{\Break}$
  or~$\isblocked(p,n)$.\footnote{We sometimes abbreviate ``$p=\ceu{\Nop}$
    or~$p=\ceu{\Break}$'' as~``$p=\ceu{\Nop,\Break}$''.}
\end{definition}

@@

The syntactic restriction discussed in Section~\ref{sec.ceu.evts} regarding
the body of loops and the restriction mentioned in
Section~\ref{sec.sem.opsem} about the body of~$\ceu{\Fin}$ statements are
formalized in Assumption~\ref{ass.x.syn-rest} below.  These restrictions are
essential to prove the next theorem.

\begin{assumption}[Syntactic restrictions]\strut
  \label{ass.x.syn-rest}
  %%
  \begin{enumerate:a}
  \item\label{ass.x.syn-rest.fin} If~$p=\ceu{\Fin{p_1}}$ then~$p_1$ contains
    no occurrences of statements~$\ceu{\AwaitExt}$, $\ceu{\AwaitInt}$,
    $\ceu{\Every}$, or~$\ceu{\Fin}$.  And so, for any~$n$,
    $\<\clear(p_1),n,\nil>\nst[*]\<\ceu{\Nop},n,\nil>$.
    %%
    % \[
    %   \<\clear(p_1),n,\nil>\nst[*]\<\ceu{\Nop},n,\nil>\,.
    % \]
    %%
  \item\label{ass.x.syn-rest.loop} If~$p=\ceu{\Loop{p_1}}$ then all
    execution paths of~$p_1$ contain a matching~$\ceu{\Break}$ or
    an~$\ceu{\AwaitExt}$.  Consequently, for all~$n$, there are~$p_1'$
    and~$e$ such that
    $\<\ceu{\Loop{p_1}},n,\nil>\nst[*]\<p_1',n,e>$,
    %%
    % \[
    %   \<\ceu{\Loop{p_1}},n,\nil>\nst[*]\<p_1',n,e>\,,
    % \]
    %%
    where $p_1'=\ceu{{\Break}\AtLoop{p_1}}$ or~$\isblocked(p_1',n)$\,.
  \end{enumerate:a}
\end{assumption}

@@

The main result of this section, Theorem~\ref{thm.x.term}, is similar to
Theorem~\ref{thm.x.term-nst-*} but applies to transitions~$\trans$ in
general.  Before stating and proving it, we need to characterize irreducible
descriptions in general.  This characterization, Definition~\ref{def.x.H},
and depends on the notions of potency and rank.

\begin{definition}
  \label{def.x.pot}
  %%
  The \emph{potency} of a program~$p$ in reaction to event~$e$,
  denoted~$\pot(p,e)$, is the maximum number of~$\ceu{\EmitInt}$ statements
  that can be executed in a reaction of~$p$ to~$e$, i.e.,
  \[
    \pot(p,e)=\pot'(\bcast(p,e))\,,
  \]
  where~$\pot'$ is an auxiliary function that counts the maximum number of
  reachable~$\ceu{\EmitInt}$ expressions in the program resulting from the
  broadcast of event~$e$ to~$p$.

  The function~$\pot'$ is defined by the following clauses:
  \begin{enumerate:a}
  \item\label{def.x.pot.first}$\pot'(\ceu{\EmitInt}(e))=1$.
    %%
  \item$\pot'(\ceu{\IfElse{\Mem(\Id)}{p_1}{p_2}})=$\\
    \strut\quad$\max\{\pot'(p_1),\pot'(p_2)\}$.
    %%
  \item$\pot'(\ceu{\Loop{p_1}})=\pot'(p_1)$.
    %%
  \item$\pot'(\ceu{{p_1}\And{p_2}})=\pot'(p_1)+\pot'(p_2)$.
    %%
  \item$\pot'(\ceu{{p_1}\Or{p_2}})=\pot'(p_1)+\pot'(p_2)$.
    %%
  \item If~$p_1\ne\ceu{\Break},\ceu{\AwaitExt(e)}$,
    \begin{align*}
      \pot'(\ceu{p_1;\,p_2})&=\pot'(p_1)+\pot'(p_2)\\
      \pot'(\ceu{p_1\AtLoop{p_2}})&=
      \begin{cases}
        \pot'(p_1)              &\text{if\enspace(\dag)}\\
        \pot'(p_1)+\pot'(p_2)   &\text{otherwise,}\\
      \end{cases}
    \end{align*}
    where~(\dag) stands for: ``a~$\ceu{\Break}$ or~$\ceu{\AwaitExt}$ occurs
    in all execution paths of~$p_1$''.
    %%
  \item If~$p_1,p_2\ne\ceu{\Break}$,
    \[
      \pot'(\ceu{{p_1}\AtAnd{p_2}})=\pot'(p_1)+\pot'(p_2)\,.
    \]
    %%
  \item\label{def.x.pot.before-last} If~$p_1,p_2\ne\ceu{\Break}$
    and~$p_1,p_2\ne\ceu{\Nop}$,
    \[
      \pot'(\ceu{{p_1}\AtOr{p_2}})=\pot'(p_1)+\pot'(p_2)\,.
    \]
    %%
  \item\label{def.x.pot.last} Otherwise, if none
    of~\ref{def.x.pot.first}--\ref{def.x.pot.before-last} applies,
    $\pot(\_)=0$.
  \end{enumerate:a}
\end{definition}

@@

Intuitively, the rank of a description~$\delta$ is a measure of the maximum
amount of ``work'' (transitions) required to transform~$\delta$ into an
irreducible description, as defined bellow.

@@

The next two lemmas establish that a single application of~$\out$ or~$\nst$
either preserves or decreases the rank of the input description.  All rank
comparisons assume lexicographic order, i.e., if~$\rank(\delta)=\<i,j>$
and~$\rank(\delta')=\<i',j'>$ then~$\rank(\delta)>\rank(\delta')$ iff~$i>i'$
or~$i=i$ and~$j>j'$.
%%
The proof of Lemma~\ref{lem.x.rank-out} follows directly from~\R{push}
and~\R{pop} and from Definitions~\ref{def.x.pot} and~\ref{def.x.rank}.  The
proof of Lemma~\ref{lem.x.rank-nst}, however, is by induction on the
structure of~$\nst$ derivations.  Since Lemma~\ref{lem.x.rank-nst} is of
fundamental importance for the termination result, we present a sketch of its
proof.
